|

### 2021-09-16 paper.Rmd
writing

<!-- % still, we know little about the role of discrete emotions in interacting with misinformation and corrective information -->
While this has been linked to affect [@thorson2016belief], the connection of distinct emotions to both susceptibility to misinformation, and the effectiveness of corrective information is under-explored. This is unfortunate given the important role that emotional appeals play in the dissemination of false or misleading information. We know that false news spread faster than true news, possibly because they are more likely to evoke emotions of surprise and disgust [@vosoughi2018spread]. In general, moral-emotional language is rewarded on social media [@brady2017emotion].
<!-- and populist parties use more negative emotional appeals in their communication than non-populist parties [@widmann2021how]. -->

@@

* Definition of related terms: misinformation, disinformation, fake news, misperceptions [@flynn2017nature; @jerit2020political; @kuklinski2000misinformation].
* Wide-spread misperceptions may distort public opinion and corrective information (fact-checking) is one way to tackle the problem.
* People may resist corrective information due to directional motivated reasoning, sometimes even leading to "backfire" effects [@nyhan2010corrections] where misperceptions are more strongly held after correction, but most of the time corrections are at least partially successful [@walter2020fact; @wood2019elusive].
* However, while corrective information usually moves factual beliefs in the appropriate direction, the effects on other outcomes such as attitudes, blame attributions, or evaluations are mixed [@bisgaard2019getting; @nyhan2019taking; @swire2019they]
* An especially troubling case are "belief echoes", where effects of misinformation on attitudes persist despite successful correction of the factual belief [@thorson2016belief].
* Although we know that emotions may contribute to the spread of false information [@vosoughi2018spread], there is little evidence on the role of emotions for misinformation and corrective information (exceptions include [@ecker2011terrorists; @martel2020reliance; @rosenzweig2021happiness; @weeks2015emotions]).

@@

* Emotional framing is an important feature of political communication, and theories such as cognitive appraisal theory [@moors2013appraisal] or affective intelligence theory [@marcus2000affective; @marcus2019applying] provide frameworks for empirical expectaitoins about specific emotions.
* Findings from moral psychology on moralization through disgust.
* Importance of moral conviction for compromise, openness to scientific findings on consequences versus absolutist opposition.
* Of course, people are free to have moral convictions about any issue, but if emotional frames can induce moral conviction, we might have to worry about emotionally framed **mis**information also inducing moral conviction.

<!-- Although there is a rich literature on emotions in politics shedding light on its effect on information processing, political learning and participation, the connection of distinct emotions to both susceptibility to misinformation and the effectiveness of corrective information is under-explored. -->

@@

To investigate the direct effects of misinformation and corrective information conditional on emotional framing, I conducted a pre-registered survey experiment.^[The pre-registration can be found on OSF: https://osf.io/y63xu.] The experiment was implemented within wave 53 of the German Internet Panel (GIP) [@blom2015setting]. The GIP is a longitudinal panel survey based on a probability sample of the general population in Germany. Fieldwork for this study took place in May 2021 and the sample size reached 4,260 respondents.

@@

First, \@ref(fig:misinfo-only) reports the effects of receiving only the misinformation message. The upper half of the figure shows two plots with the effects of the low emotion misinformation and the lower half shows two plots with the effects of the high emtoion misinformation message. The y-axis indicates the different outcome variables, and the x-axis indicates the effect size relative to the control group, which serves as the reference category for all models.

In the top row of the lower right panel, we see the largest effect, which results from the high emotional frame on the factual belief outcome. Respondents receiving the high emotional misinformation about GM food agree to the statement that GM food is safe to eat with a score that is 16 percentage points less than the control group. For the low emotion misinformation, the effect is 13 percentage points, indicating that both messages are persuasive in bringing the average factual belief farther from the scientific consensus belief.

Concerning effects on emotions, policy opinion and moral conviction, results largely replicate results by @clifford2019emotional.

@@

After having established that a persuasive, emotionally framed message containing misinformation can influence policy opinion and increase moral conviction, I now report the results of the neutral correction in figure \@ref(fig:neutral-correction).

* the neutral correction is effective for the low emotion misinformation, but not the high emotion misinformation

```{r neutral-correction, fig.align="center", fig.cap="Effects of the neutral correction for misinformation with low and high emotional framing Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}

@@

Finally, results from the counterframe message show that a correction embedded in a persuasive positive frame is successful in bringing all outcome measures to the control group level (see figure \@ref(fig:counterframe). 
It should be noted, however, that there is also no positive effect of the positively framed message on beliefs and opinions towards GM food, although the counterframe message itself is persuasive (figure \@ref(fig:info-only) in the appendix).

```{r counterframe, fig.align="center", fig.cap="Effects of the counterframe correction for misinformation with low and high emotional framing. Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}

@@

```{r info-only, fig.align="center", fig.cap="Effects of the neutral correction and counterframe correction without previous misinformation. Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}

>>>>

<!-- % still, we know little about the role of discrete emotions in interacting with misinformation and corrective information -->
While this has been linked to affect [@thorson2016belief], the connection of distinct emotions to both susceptibility to misinformation, and the effectiveness of corrective information is under-explored (exceptions include [@ecker2011terrorists; @martel2020reliance; @rosenzweig2021happiness; @weeks2015emotions]. This is unfortunate given the important role that emotional appeals play in the dissemination of false or misleading information. We know that false news spread faster than true news, possibly because they are more likely to evoke emotions of surprise and disgust [@vosoughi2018spread]. In general, moral-emotional language is rewarded on social media [@brady2017emotion].
<!-- and populist parties use more negative emotional appeals in their communication than non-populist parties [@widmann2021how]. -->

@@

* Definition of related terms: misinformation, disinformation, fake news, misperceptions [@flynn2017nature; @jerit2020political; @kuklinski2000misinformation].
* Wide-spread misperceptions may distort public opinion and corrective information (fact-checking) is one way to tackle the problem.
* People may resist corrective information due to directional motivated reasoning, sometimes even leading to "backfire" effects [@nyhan2010corrections] where misperceptions are more strongly held after correction, but most of the time corrections are at least partially successful [@walter2020fact; @wood2019elusive].
* However, while corrective information usually moves factual beliefs in the appropriate direction, the effects on other outcomes such as attitudes, blame attributions, or evaluations are mixed [@bisgaard2019getting; @nyhan2019taking; @swire2019they]
* An especially troubling case are "belief echoes", where effects of misinformation on attitudes persist despite successful correction of the factual belief [@thorson2016belief].
* Although we know that emotions may contribute to the spread of false information [@vosoughi2018spread], there is little evidence on the role of emotions for misinformation and corrective information.

@@

* Emotional framing is an important feature of political communication, and theories such as cognitive appraisal theory [@moors2013appraisal] or affective intelligence theory [@marcus2000affective; @marcus2019applying] provide frameworks for empirical expectaitoins about specific emotions.

* Findings from moral psychology on moralization through disgust.
* Importance of moral conviction for compromise, openness to scientific findings on consequences versus absolutist opposition.

Although there is a rich literature on emotions in politics shedding light on its effect on information processing, political learning and participation, the connection of distinct emotions to both susceptibility to misinformation and the effectiveness of corrective information is under-explored.

A study has shown that moral shock works to moralize attitudes towards GM food.

However, the main part of the material used there can be classified as misinformation
GM food is an issue on which disinformation exists. This does not mean that there can or should be no criticism, but concrete adverse health effects are not warranted by evidence. Least knowledgeable people are the fiercest opponents. So

@@

To investigate the direct effects of misinformation and corrective information conditional on emotional framing, I conducted a pre-registered survey experiment.^[Pre-registration can be found on OSF: https://osf.io/y63xu.] The experiment was implemented within wave 53 of the German Internet Panel (GIP) [@blom2015setting]. The GIP is a longitudinal panel survey based on a probability sample of the general population in Germany. Fieldwork for this study took place in May 2021 and the sample size reached 4,260 respondents.

@@

First, \@ref(fig:misinfo-only) reports the effects of reiceiving only the misinformation message. The upper half of the figure shows two plots with the effects of the low emotion misinformation and the lower half shows two plots with the effects of the high emtoion misinformation message. The y-axis indicates the different outomce variables, and the x-axis indicates the effect size relative to the control group, which serves as the reference category for all models.

@@

```{r fig-2, fig.align="center", fig.cap="Effects of the neutral correction for misinformation with low and high emotional framing Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}

@@

```{r fig-3, fig.align="center", fig.cap="Effects of the counterframe correction for misinformation with low and high emotional framing. Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}

@@

```{r fig-4, fig.align="center", fig.cap="Effects of the neutral correction and counterframe correction without previous misinformation. Coefficients (and 95 confidence intervals) from OLS-regressions with the outcome variable rescaled from 0 to 1 and the control group as the reference category.", fig.pos="H", fig.width=8, fig.height=6, message=FALSE, warning=FALSE, echo = FALSE, paged.print=FALSE}
